---
title: 'CMap Clues for drug-target illumination'
author: "Jeremy Yang"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
---

# Introduction

LINCS is an NIH initiative, generally a superset of CMAP, a Broad Institute project. Clue.io is the software
platform for CMAP data exploration. The Clue.io API supplied the data for this workflow.

Important terms:

* `Landmark genes`: L1000 set, selected for experiments for optimal genomic representation.
* `Best inferred`, `inferred`, `not inferred`: classes of non-landmarks relative to landmarks.
* `Cell lines`: used in experiments; associated with diseases, tissues and organs.
* `Perturbagens`: small molecules or genetic.
* `L1000 Assay`: measures effect of perturbagen against landmarks, up/down regulated genes.
* `Profiles`: of perturbagens, from assays including L1000.
* `Signatures`: aggregated profiles(?)

See also: <https://clue.io/connectopedia/category/Analytical%20Methods>

## Applying CMAP for drug-target illumination

In general, genes which are regulated by a perturbagen are more likely to be associated with tissues and
diseases also associated with that perturbagen. The signatures of perturbagens are polypharmalogical 
descriptors which could suggest repurposing hypotheses, e.g. where drugs with different indications have 
similar profiles, new indications are suggested.

__However, at this time it is not clear to me (JJY) what is a strong direction for drug-target hypothesis generation,
by integrating CMAP and GTEx/Ex-files data.__


```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r echo=FALSE, message=FALSE}
library(readr)
library(data.table)
library(plotly, quietly=T)
```

# Genes:

```{r}
genes <- read_delim("data/cmap_genes_all.tsv", "\t", col_types = cols(entrez_id = col_character()))
setDT(genes)
sprintf("genes: %d", nrow(genes))
```

## L1000 Landmark and Inferred Types
```{r}
l1000_types <- genes[, .(.N), by="l1000_type"]
p1 <- plot_ly(data=l1000_types, type="pie", labels=~l1000_type, values=~N, textinfo="label+value+percent", textposition="auto") %>%
  layout(title="Genes by L1000 type", margin=list(t=100, b=60), font=list(family="Monospace", size=20), showlegend=F)
p1
```

# Perturbagens (small molecules, not genetic)
```{r message=FALSE}
perts <- read_delim("data/cmap_perts_all.tsv", "\t", col_types = cols(pubchem_cid = col_character()))
setDT(perts)
sprintf("perturbagens: %d ; compounds (pert_type = trt_cp): %d", nrow(perts), sum(perts$pert_type=="trt_cp"))
perts <- perts[pert_type=="trt_cp"]
```

## Perturbagen classes 
Available for some compounds (drugs?).
```{r message=FALSE}
pcls <- read_delim("data/cmap_pcls.tsv", "\t")
setDT(pcls)
pert2pcl <- perts[!is.na(pcl_membership), .(pert_id, pcl_membership)]
pert2pcl <- pert2pcl[, lapply(.SD, function(x) unlist(tstrsplit(x, ";", fixed=TRUE))), by = pert_id][!is.na(pcl_membership)]
setnames(pert2pcl, old="pcl_membership", new="pcl")
pert2pcl <- merge(pert2pcl, pcls, by.x="pcl", by.y="group_id", all.x=T, all.y=F)
knitr::kable(pert2pcl[, .(N_perts = .N), by="pcl"][order(-N_perts)][1:10], caption="PCLs (top 10)")
```

```{r}
p2 <- plot_ly(x=perts$num_sig, type="histogram") %>%
  layout(title=paste0("Clue perturbagens (compounds), N_signatures<br>(N_total = ", nrow(perts), ")"), 
         xaxis=list(type="log", title="N_signatures"), yaxis=list(title="N"),
         margin=list(t=120), font=list(family="Arial", size=20), showlegend=F)
p2
```
#

# Cell Lines:
For known primary disease only.
```{r message=FALSE}
cells <- read_delim("data/cmap_cells_all.tsv", "\t")
setDT(cells)
n_cells_total <- nrow(cells)
cells <- cells[!grepl("normal", primary_disease) & !grepl(";", primary_disease) & !(is.null(primary_disease)) & primary_disease!="other"]
sprintf("Cell lines: %d ; with known primary disease: %d", n_cells_total, nrow(cells))
```

## Primary diseases:
```{r}
diseases <- cells[, .(N_cells = .N), by="primary_disease"][order(-N_cells)]
sprintf("Cell line primary_disease count: %d", length(unique(diseases$primary_disease)))
knitr::kable(diseases[1:10], caption="Cell Lines count by diseases (top 10)")
p3 <- plot_ly(data=diseases, type="pie", labels=~primary_disease, values=~N_cells, textinfo="label", textposition="auto") %>%
  layout(title=sprintf("Clue cell lines by primary_disease<br>N_cells = %d ; N_disease = %d", nrow(cells), nrow(diseases)), margin=list(t=100, b=60), font=list(family="Monospace", size=12), showlegend=T)
p3
```

# Profiles
Assay-well level data.

Fields:

* assay
* cell_id
* pert_id, pert_desc
* det_plate, det_well

### Read sample profiles for psychiatric drug aripiprazole

```{r message=FALSE}
profiles <- read_delim("data/cmap_profiles_aripiprazole.tsv", "\t")
sprintf("profiles: %d ; perts: %d ; assays: %d ; cells: %d", nrow(profiles), length(unique(profiles$pert_id)),
        length(unique(profiles$assay)), length(unique(profiles$cell_id)))
```

# Signatures
For a given cell_id and pert_id, there may be a signature, which will indicate up- and down- regulated genes.
(Bing = Best inferred genes, includes Landmark)

Fields:

* pert_id, pert_iname, pert_desc, pert_dose
* cell_id, provenance_code
* target_is_lm, target_is_bing, target_zs
* dn100_bing: 100 most downregulated genes 
* up100_bing: 100 most upregulated genes

### Read sample signatures for psychiatric drug aripiprazole
Remove BING genes with ambiguous UP/DOWN regulation.

```{r message=FALSE}
sigs <- read_delim("data/cmap_sigs_aripiprazole.tsv", "\t")
setDT(sigs)
sprintf("sigs: %d ; perts: %d ; cells: %d", nrow(sigs), length(unique(sigs$pert_id)),
        length(unique(sigs$cell_id)))
sig2upg <- sigs[, .(pert_id, bing = up100_bing)]
sig2upg <- sig2upg[, lapply(.SD, function(x) unlist(tstrsplit(x, ";", fixed=TRUE))), by=pert_id][!is.na(bing)]
sig2upg <- unique(sig2upg)
sig2upg <- sig2upg[order(as.integer(sig2upg$bing))]
sig2upg[["up"]] <- TRUE

sig2dng <- sigs[, .(pert_id, bing = dn100_bing)]
sig2dng <- sig2dng[, lapply(.SD, function(x) unlist(tstrsplit(x, ";", fixed=TRUE))), by=pert_id][!is.na(bing)]
sig2dng <- unique(sig2dng)
sig2dng <- sig2dng[order(as.integer(sig2dng$bing))]
sig2dng[["down"]] <- TRUE
sig2bing <- merge(sig2upg, sig2dng, by=c("pert_id", "bing"), all=T)
sig2bing <- sig2bing[order(as.integer(sig2bing$bing))]
sprintf("UP/DOWN-regulated genes, up: %d ; down: %d ; up_AND_down (removed): %d", 
        nrow(sig2bing[(up)]), nrow(sig2bing[(down)]), nrow(sig2bing[(up)&(down)]))
sig2bing <- sig2bing[(is.na(up) | is.na(down))]
```

### Gene symbols, names etc. from TCRD

```{r message=FALSE}
tcrd <- read_delim("/home/data/TCRD/data/tcrd_targets.tsv", "\t",
                       col_types = cols_only(fam = col_character(), name = col_character(),
                                             tdl = col_character(), geneid = col_character(),
                                             sym=col_character(), uniprot=col_character(),
                                             stringid=col_character()))
setDT(tcrd)
#setnames(tcrd, old=names(tcrd), new=sub("^.*\\.", "", names(tcrd)))
tcrd$fam[tcrd$fam=="None"] <- NA
sig2bing <- merge(sig2bing, tcrd, by.x="bing", by.y="geneid", all.x=T, all.y=F)
```

### Up/down regulated genes by TDL and family:

```{r message=FALSE}
tdl_counts <- sig2bing[!is.na(tdl), .(N_genes = .N), by=c("fam", "tdl")]
tdls <- c("Tdark", "Tbio", "Tchem", "Tclin")
tdl_counts[["tdl"]] <- factor(tdl_counts$tdl, levels=tdls)
tdl_counts <- tdl_counts[order(fam, tdl)]
#
p4 <- plot_ly(data=tdl_counts, type="bar", x=~fam, y=~N_genes, color=~tdl, colors=c("#333333","red","green","blue"),
	text=tdl_counts$N_genes, textposition = 'auto') %>%
	layout(title=sprintf("+/- regulated genes for aripiprazole<br>N_total = %d ; N_fam_none = %d", nrow(sig2bing), nrow(sig2bing[is.na(fam)])), 
	barmode="group", showlegend=T, legend=list(x=.8, y=.9),
	margin=list(t=120, b=60, r=60), font=list(family="Monospace", size=16))
p4
#knitr::kable(tdl_counts, caption="+/- regulated genes by TDL and family")
```

### GPCRs:
```{r}
knitr::kable(sig2bing[fam=="GPCR", .(pert_id, bing, up_or_down = ifelse((up), "UP", ifelse((down), "DOWN", "?")), sym, name, tdl)], caption="+/- regulated GPCRs")
```


