---
title: 'CMap Clues for disease drug-target associations'
author: "Jeremy Yang"
output:
  html_document:
    number_sections: yes
    toc: yes
---

For explanation of "inferred", etc., see https://clue.io/connectopedia/category/Analytical%20Methods

```{r echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r setup, echo=TRUE, message=FALSE}
library(readr)
library(data.table)
library(plotly, quietly=T)
```

# Genes:
```{r message=FALSE}
genes <- read_delim("data/cmap_genes_all.tsv", "\t", col_types = cols(entrez_id = col_character()))
setDT(genes)
sprintf("genes: %d", nrow(genes))
```

## L1000 Landmark and Inferred Types
```{r}
l1000_types <- genes[, .(.N), by="l1000_type"]
p1 <- plot_ly(data=l1000_types, type="pie", labels=~l1000_type, values=~N, textinfo="label+value+percent", textposition="auto") %>%
  layout(title="Genes by L1000 type", margin=list(t=100, b=60), font=list(family="Monospace", size=20), showlegend=F)
p1
```

# Perturbagens (small molecules, not genetic)
```{r message=FALSE}
perts <- read_delim("data/cmap_perts_all.tsv", "\t", col_types = cols(pubchem_cid = col_character()))
setDT(perts)
sprintf("perturbagens: %d ; compounds (pert_type = trt_cp): %d", nrow(perts), sum(perts$pert_type=="trt_cp"))
perts <- perts[pert_type=="trt_cp"]
```

## Perturbagen classes 
Available for some compounds (drugs?).
```{r message=FALSE}
pcls <- read_delim("data/cmap_pcls.tsv", "\t")
setDT(pcls)
pert2pcl <- perts[!is.na(pcl_membership), .(pert_id, pcl_membership)]
pert2pcl <- pert2pcl[, lapply(.SD, function(x) unlist(tstrsplit(x, ";", fixed=TRUE))), by = pert_id][!is.na(pcl_membership)]
setnames(pert2pcl, old="pcl_membership", new="pcl")
pert2pcl <- merge(pert2pcl, pcls, by.x="pcl", by.y="group_id", all.x=T, all.y=F)
knitr::kable(pert2pcl[, .(N_perts = .N), by="pcl"][order(-N_perts)][1:10], caption="PCLs (top 10)")
```

```{r}
p3 <- plot_ly(x=perts$num_sig, type="histogram") %>%
  layout(title=paste0("Clue perturbagens (compounds), N_signatures<br>(N_total = ", nrow(perts), ")"), 
         xaxis=list(type="log", title="N_signatures"), yaxis=list(title="N"),
         margin=list(t=120), font=list(family="Arial", size=20), showlegend=F)
p3
```
#

# Cell Lines:
For known primary disease only.
```{r message=FALSE}
cells <- read_delim("data/cmap_cells_all.tsv", "\t")
setDT(cells)
n_cells_total <- nrow(cells)
cells <- cells[!grepl("normal", primary_disease) & !grepl(";", primary_disease) & !(is.null(primary_disease)) & primary_disease!="other"]
sprintf("Cell lines: %d ; with known primary disease: %d", n_cells_total, nrow(cells))
```

## Primary diseases:
```{r}
diseases <- cells[, .(N_cells = .N), by="primary_disease"][order(-N_cells)]
sprintf("Cell line primary_disease count: %d", length(unique(diseases$primary_disease)))
knitr::kable(diseases[1:10], caption="Cell Lines count by diseases (top 10)")
p4 <- plot_ly(data=diseases, type="pie", labels=~primary_disease, values=~N_cells, textinfo="label", textposition="auto") %>%
  layout(title=sprintf("Clue cell lines by primary_disease<br>N_cells = %d ; N_disease = %d", nrow(cells), nrow(diseases)), margin=list(t=100, b=60), font=list(family="Monospace", size=12), showlegend=T)
p4
```

# Profiles
Assay-well level data.

Fields:
* assay
* cell_id
* pert_id, pert_desc
* det_plate, det_well

# Signatures
For a given cell_id and pert_id, there may be a signature, which will indicate up- and down- regulated genes.
(Bing = Best inferred genes, includes Landmark)

Fields:
* pert_id, pert_iname, pert_desc, pert_dose
* cell_id, provenance_code
* target_is_lm, target_is_bing, target_zs
* dn100_bing: 100 most downregulated genes 
* up100_bing: 100 most upregulated genes